Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-06-06 manuscript/311 manuscript.tex
Weekly updates

	\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
	An additional  problem discovered with the created\_date and closed\_date fields; there appears to be an unusually large number of 
	SRs created or closed at exactly midnight (00:00:00) and noon (12:00:00), to the second. The distribution of SR creation and closure 
	largely follows the work-day clock with lots of SRs created during day-light hours, and fewer SRs 
	created in the early hours of the morning. However, here there appears to be significantly greater numbers of SR closed exactly at midnight and exactly at noon,
	as well as a significant number of SRs created exactly at noon. The reason for this anomaly is worth investigating 
	as it affects the SR ``duration'' question; if SR durations are used to measure responsiveness
	for the delivery of services, and if those times are not an accurate reflection of the start/finish of the Service, then the use of
	duration as a service quality metric is called into question. Observations during this 2-year period include:

	\begin{itemize}
		    \item There were 235,347 SRs closed exactly at midnight
		    \item There were 105,505 SRs closed exactly at noon. 
		    \item There were 99,779 SRs created exactly at noon 
	\end{itemize}

	Here is a visualization of SRs closed and created by the hour of the day :

@@

	The closed\_date chart show significant spikes in the number of SRs at both midnight
	and noon with significant counts at exactly those times, to the second. The midnight spike is certainly suspect compared to the hours just before and after midnight where
	there is normally a pronounced slump in SR closures. And both the closed-at-noon and closed-at-midnight are 
	approaching 2\textsigma's from the hourly average. The created\_date shows a maximum value at noon also with a significant count occurring exactly at noon, again to 
	the second.

	These unusual patterns of created \& closed SRs at exactly the hours of midnight and noon perhaps indicates
	the presence of a bulk process; one that perhaps automatically ``closes'' (or ``creates'') a large number of SRs with a provided
	time-stamp of midnight (00:00:00) or noon (12:00:00). If so, then addressing this issue likely requires a closer look at those 
	Agencies providing these suspect SR closed/created times. Clearly the desirable state is that the created\_date and closed\_date fields
	reflect the true opening and closing of the SRs, such that an accurate ``duration'' can be computed which accurately
	represents the SR life-cycle.

	Looking at the distribution by Agency for the ``closed-exactly-at-midnight'' shows that \textgreater{}90\% of these suspect SRs come from just two Agencies - Dept. of 
	Buildings (DOB) and Dept. of Sanitation (DSNY). If you include HPD and DOT, that accounts for 99\% of the suspect SR closures.

@@

	Hoever, the ``SR closed at noon'' visualization shows that  a single Agency, DSNY,
	is responsible for \textgreater{}99\% of the SRs ``closed-exactly-at-noon''. Accordingly, any effort to resolve this issue
	necessitates working with DSNY to fully understand what is occurring.

@@

	It is difficult to determine if these observations are indeed anomalies. They could be the result of normal, expected
	311 SR activity. But at least by observation that appears unlikely.

>>>>

	\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
	An additional problem discovered with the created\_date and closed\_date fields; there appears to be an unusually large number of 
	SRs created or closed at exactly midnight (00:00:00) and exactly noon (12:00:00), to the second. The distribution of SR creation and closure 
	largely follows the work-day clock with many SRs created during day-light hours, and fewer SRs 
	created at night and in the early hours of the morning. 
	However, here there appears to be significantly greater numbers of SR closed exactly at midnight and exactly at noon,
	as well as a significant number of SRs created exactly at noon. The reason behind this anomaly is worth investigating 
	as it affects the SR ``duration'' question. If SR durations are used to measure responsiveness
	for the delivery of services, and if those times are not an accurate reflection of the start/finish of the Service, then the use of
	duration as a service quality metric is called into question. Observations during this 2-year period include:

	\begin{itemize}
		    \item There were 99,779 SRs created exactly at noon (12:00:00)
		    \item There were 235,347 SRs closed exactly at midnight (00:00:00)
		    \item There were 105,505 SRs closed exactly at noon. 
	\end{itemize}

	There is, of course, a broad distribution of the time that SRs are created by day and hour, by night and day, by weekday and weekend. 
	One would think that that distribution would be (mostly) evenly distributed by the minute and second. That is, 
	if there are normally 100 SRs created between 1-2pm, that the distribution to the minute and second 
	would be close to random. For example, there is no reason to suspect that a larger number of SRs would be 
	created at 6 minutes past the hour rather than at 5 minutes past the hour. And certainly, if you look at the second that these SRs
	are created you would again expect a (more-or-less) random distribution; that there would be no reason to believe 
	that a higher (or lower) number of SRs would arrive at 18 seconds past the minute versus at 39 seconds past the minute.
	Unfortunately, that turned out not to be true.

	Let's look at the SR creation distribution. We discovered that there are a significant number of SRs created exactly at noon. To reveal this
	trend, we looked at the exact date-time stamp of the created\_date(s). We began by selected only SRs that were created exactly on the minute;
	at 09:00:00 or 13:00:00 for example.  Here is a visualization of the number of SRs created and closed by hour-of-the-day:

@@

	The created\_date shows a maximum value at 12:00. The closed\_date chart show significant spikes at both midnight
	and noon. The midnight (00:00) spike appears to be an anomaly especially as compared to the hours just 
	before and after midnight where there is normally a pronounced slump in SR closures. 
	And both the closed-at-noon and closed-at-midnight are approaching 2\textsigma's from the hourly average.

	Let's drill down on each of these outlier points. First, we aggregate the created\_date over the 2-year period that fall exactly
	on the top-of-the-hour with the minute and second values equal to 0, e.g. 09:00:00, 14:00:00. This visualization
	reveals a clearer picture of where these spikes are:

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{SR_created_by_top_of_hour.png}
		 \caption{SRs Created at Top of the Hour}
		 \label{fig:tophourcreated}
	\end{figure}

	One other way to visualize this anomaly is to look at the busiest day during this 2-year period (Friday, 2023-09-29).
	Here we will aggregated the created\_date(s) by minute (with seconds equal to zero). This provides a 
	minute-by-minute look at SR creation on the busiest day of this study. Here we see a clear spike at 
	exactly noon (12:00), well beyond the 3\textsigma line, and well above any other hour of the day.

	\begin{figure}[H]
		 \centering
		 \includegraphics[width=\textwidth]{SR_created_by_minute_of_busiest_day.png}
		 \caption{SRs created minute-by-minute on busiest day}
		 \label{fig:busiestcreated}
	\end{figure}

	Let's take a similar approach to the closed\_date distribution over the 2-year period. We aggregate 
	SRs by close\_date to find the top-of-the-hour, where minute and second values equal to 0, e.g. 20:00:00. 
	Here we see spikes at both midnight (00:00) and at noon (12:00), with the midnight spike being 
	well beyond the 3\textsigma line. Also note how much lower (and nearly equal) all the other
	top-of-the-hour values are, relative to the midnight and noon spikes.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{SR_closed_by_top_of_hour.png}
		 \caption{SRs Closed at Top of the Hour}
		 \label{fig:tophourclosed}
	\end{figure}

	As before, let's visualize this anomaly by looking at the closed\_date(s) on the busiest day
	during this 2-year study period (Friday, 2023-09-29). We aggregate closed\_date(s) by the
	minute (with seconds equal to zero). This gives a minute-by-minute view of SR closure
	on the busiest day of this study. Clearly visual is the very large closure spikes occurring
	at both midnight and noon. The chart highlights both of these anomalies with spikes
	of 286 and 120 at 00:00:00 and 12:00:00 respectively. Again, note how these spikes
	are so very much above the median values for all other minutes.

 	\begin{figure}[H]
		 \centering
		 \includegraphics[width=\textwidth]{SR_closed_by_minute_of_busiest_day.png}
		 \caption{SRs closed minute-by-minute on busiest day}
		 \label{fig:busiestclosed}
	\end{figure}

	These unusual patterns of created \& closed SRs at exactly the hours of midnight and noon likely indicates
	the presence of a bulk create/close software process; a process that perhaps automatically ``closes'' (or ``creates'') 
	a large number of SRs with a provided time-stamp of midnight (00:00:00) or noon (12:00:00), and does so in bulk. 
	If so, then addressing this issue likely requires a closer look at those Agencies providing these suspect SR closed/created times.
	 The desirable state is that the created\_date and closed\_date fields correctly reflect the true opening and closing 
	 of the SRs, such that an accurate ``duration'' can be computed for the SR life-cycle. Otherwise, we are likely seeing 
	 durations that are artificially manipulated by a software process performing bulk uploads to the 311 system.

	Looking at the distribution by Agency for the ``closed-exactly-at-midnight'' shows that \textgreater{}90\% of these suspect SRs come from just two Agencies - Dept. of 
	Buildings (DOB) and Dept. of Sanitation (DSNY). If you include HPD and DOT, that accounts for 99\% of the suspect SR closures. This Agency
	distribution does not follow the overall SR distribution and we can assume that it is Agency specific.

@@

	However, the ``SR closed at noon'' visualization shows that  a single Agency, DSNY,
	is responsible for \textgreater{}99\% of the SRs ``closed-exactly-at-noon''. Accordingly, any effort to resolve this issue
	necessitates working with DSNY to fully understand how this is occurring.

@@

	While it is difficult to ultimately know if these observations are indeed anomalies; they could be the result of normal, expected
	311 SR activity. But at least by observation that outcome appears very unlikely.
